\documentclass{article}
%
% BASIC PACKAGES
\usepackage{amsmath,amssymb,amsthm,enumitem} % Some standard math packages.
\usepackage{titling} % Enables \setlength{\droptitle}
\usepackage{parskip} % Cleaner paragraph display
\usepackage[margin=0.75in]{geometry} % Adjusts margins.
\usepackage[utf8]{inputenc} % Use UTF-8 input encoding instead of default ASCII.
\usepackage{fancyvrb} % Allows Verbatim sections with line numbers and such. Note the capital V.
\usepackage{xcolor} % for text color, e.g. \todo command
\usepackage{ragged2e} % For text alignment environments, e.g. \begin{center}
\usepackage{enumitem} % Allows different list enumerations, like a.) b.) c.)
\usepackage{nameref} % Allows using \nameref{refname} to insert links to sections by title

% HYPERLINKS
\usepackage{hyperref} % For \href and \uri
\hypersetup{
  colorlinks=true,
  urlcolor=green,
  linkcolor=green
}

% BIBLIOGRAPHY
\usepackage[english]{babel}
\usepackage[
  backend=biber,
  style=numeric,
  sorting=ynt
]{biblatex}
\addbibresource{bibliography.bib}

% CUSTOM COMMANDS
\newcommand {\todo}[1] {{\textbf{\color{red}#1}}}
\newcommand {\code}[1] {\texttt{#1}}
\newcommand {\term}[1] {\textbf{#1}}

\begin{document}

% Header
\begin{center}
    \Huge Re-implementing a Ruby on Rails Site in Yesod \\
    \large CS 557: Functional Languages @ Portland State University \\
    Dylan Laufenberg, winter 2019
\end{center}

\input{introduction.tex}

\todo{New section: Cover ``The Basics''? Regardless, need some kind of transition.}

\input{resources-and-type-safe-urls.tex}

\input{shakespearean-templates.tex}

\section{Skipped sections}

In order to maximize the value of this Yesod experiment, I skip over a number of topics that the Yesod book covers in detail. I don't touch them in my implementation except where absolutely necessary, though I may refer to them as needed. Because I'm not directly and explicitly working with them, I won't discuss them directly here. See the Yesod book (as needed) for background information on:

\begin{itemize}
  \item Widgets
  \item The Yesod typeclass
  \item Forms
  \item Sessions
\end{itemize}

\todo{Contain this knowledge in new sections, etc.}

Make new Parts: Learning the Basics and Adding Features.

More about routing:
  Yesod safely breaks routes into data types that are instances of PathPiece and requires us to define two-way conversions between Text and the PathPiece instance type. Converting from text returns a Maybe value, which allows us to guard against invalid input and guarantee that we're passing valid inputs into our route handlers.
  There are three types of path pieces: static text that must be matched exactly, single pieces (denoted by the # symbol followed by the name of a PathPiece instance type), and multi-pieces (denoted by * followed by the type) that can match any number of pieces (as long as the fromPathMultiPiece function definition is satisfied). (Where do I put my instance declarations in the scaffolded site? Hm. Need to check the Scaffolded Site chapter.)
  By default, Yesod disallows overlapping routes at compile time. To override this, add ! before a route to disable its overlap checking. In this case, the first matching route wins. (I'll need to use this.)
  The resource constructor gets arguments that match the PathPiece and PathMultiPiece types.
\section{Reflections}

\todo{Collect these thoughts into a real, formalized section in its own file.}

Reflect on Yesod from the perspective of someone who's just made it through the basics. In particular, TinyOS's observation about being hard for beginners. I feel like I have to be an expert to do almost anything. Every piece of code feels like it's written for experts. And the lack of an accessible API reference (with good documentation) so far seems to be frustrating me.

Three points from TinyOS:
\begin{enumerate}
  \item By making it harder to write bugs, you also make it harder to write code
  \item By making hard things possible, you make easy things hard
  \item By tailoring the language for expert users, you alienate newcomers and stifle your language's growth.
\end{enumerate}

% \input{p2.tex}
% \input{treeVisualizer.tex}
% \input{p3.1.tex}
% \input{p3.2.tex}

\printbibliography
\end{document}
